---
title: "Sleep Trial Analysis"
author: "J Newell"
date: "17 October 2018"
output:
  word_document: default
  html_document: default
---

A randomised controlled trial was undertaken to compare sleep deprivation (SD) over an eight week period between an intervention group using sleep therapy and a control group (i.e. usual-care). One hundred and thirty four adults were randomly assigned to either an intervention or usual-care group. The primary response variable was the improvement in sleep deprivation (averaged over a week) between baseline and follow-up at eight weeks where high values of SD represent a poorer outcome. 


```{r}
library(infer)
library(tidyverse)
library(tolerance)
```



```{r}

sleep.df <- read.csv("Sleep_trial_data.csv")
glimpse(sleep.df)
```

# Summary Statistics



```{r}

sleep.df %>% group_by(Group) %>% 
  summarize(Mean=mean(Improvement.in.Sdep), 
            SD= sd(Improvement.in.Sdep))

```

```{r}

sleep.df %>% filter(Group %in% 'Treatment') %>% 
  summarize(Treatment.Mean=mean(Improvement.in.Sdep), 
            Treatment.SD= sd(Improvement.in.Sdep))

```

```{r}

sleep.df %>% filter(Group %in% 'Control') %>% 
  summarize(Control.Mean=mean(Improvement.in.Sdep), Control.SD= sd(Improvement.in.Sdep))

```

Summarise by group using _all function

```{r}
sleep.df %>%
select(Group, Improvement.in.Sdep) %>%
group_by(Group) %>%
summarise_all(funs(mean, sd), na.rm = TRUE)

```

