---
title: "Sleep Trial Analysis"
author: "J Newell"
date: "17 October 2018"
output:
  word_document: default
  html_document: default
---

A randomised controlled trial was undertaken to compare sleep deprivation (SD) over an eight week period between an intervention group using sleep therapy and a control group (i.e. usual-care). One hundred and thirty four adults were randomly assigned to either an intervention or usual-care group. The primary response variable was the improvement in sleep deprivation (averaged over a week) between baseline and follow-up at eight weeks where high values of SD represent a poorer outcome. 


```{r}
library(infer)
library(tidyverse)
library(tolerance)
```



```{r}

sleep.df <- read.csv("Sleep_trial_data.csv")
glimpse(sleep.df)
```

# Summary Statistics

```{r}

sleep.df$Group <- recode(sleep.df$Group, Treatment = "Intervention") 


```

```{r}

sleep.df %>% group_by(Group) %>% 
  summarize(Sample_Size = n(),
            Mean=mean(SDep.at.Baseline), 
            SD= sd(SDep.at.Baseline))

```

```{r}

sleep.df %>% group_by(Group) %>% 
  summarize(Sample_Size = n(),
            Mean=mean(Improvement.in.Sdep), 
            SD= sd(Improvement.in.Sdep))

```

```{r}

sleep.df %>% filter(Group %in% 'Treatment') %>% 
  summarize(Treatment.Mean=mean(Improvement.in.Sdep), 
            Treatment.SD= sd(Improvement.in.Sdep))

```

```{r}

sleep.df %>% filter(Group %in% 'Control') %>% 
  summarize(Control.Mean=mean(Improvement.in.Sdep), Control.SD= sd(Improvement.in.Sdep))

```

Scatterplot

```{r}
ggplot(sleep.df, aes(x = SDep.at.Baseline, 
                     y = SDep.at.Follow.Up,
                     colour=Group,shape =Group)) +
        geom_point(size=2) + 
  scale_colour_manual(values = c("darkGrey", "Black")) +
  ggtitle("Scatterplot of Sleep Deprivation by Group") +
  ylab("Sleep Deprivation at Follow up") +
  xlab("Sleep Deprivation at Baseline") +
  geom_abline(slope=1, intercept=0) +
  theme(text = element_text(size = 20)) +
  theme_bw()

```


Boxplot

```{r}
ggplot(sleep.df, aes(x = Group, y = Improvement.in.Sdep)) +
        geom_boxplot() + 
  ggtitle("Boxplot of Improvement in Sleep Deprivation by Group") +
  ylab("Improvement in Sleep Deprivation") +
  xlab("") +
  geom_hline(yintercept=0, linetype="dashed",color = "green", size=1)

```

95% Confidence Interval for difference in the population mean

```{r}

t.test(Improvement.in.Sdep ~ Group, data= sleep.df)

```


# better version of t test

```{r}

sleep.df %>%
t_test(Improvement.in.Sdep ~ Group, order = c("Intervention", "Control"), 
       alternative = "two_sided",
       mu=0,
       conf_level = 0.95,
       paired=FALSE,
       var.equal=FALSE)

```




## Difference relative to baseline

```{r}

sleep.df %>% summarize(Mean=mean(SDep.at.Baseline), 
            SD= sd(SDep.at.Baseline))

```


# 95% Bootstrap CI for difference in means


```{r}

sleep.boot <- sleep.df %>%
  specify(response = Improvement.in.Sdep, explanatory = Group) %>%
  generate(reps = 1000, type = "bootstrap") %>%
  calculate(stat = "diff in means", order = c("Treatment", "Control"))

percentile_ci <- get_ci(sleep.boot)

percentile_ci

```



```{r}


sleep.boot %>% visualize(endpoints = percentile_ci, direction = "between")

```


## Tolerance Interval


##  Treatment Group

```{r}

Treatment.Improvement.in.Sdep <- sleep.df %>% filter(Group %in% 'Intervention') %>% select(Improvement.in.Sdep) %>% as.data.frame()

normtol.int(Treatment.Improvement.in.Sdep$Improvement.in.Sdep, alpha = 0.05, P = 0.95, side = 2)

```

## Controls

```{r}

Control.Improvement.in.Sdep <- sleep.df %>% filter(Group %in% 'Control') %>% select(Improvement.in.Sdep) %>% as.data.frame()

normtol.int(Control.Improvement.in.Sdep$Improvement.in.Sdep, alpha = 0.05, P = 0.95, side = 2)
```


## Permutations test

```{r}


samp.diff <- sleep.df %>%
specify(response = Improvement.in.Sdep, explanatory = Group) %>%
  calculate(stat = "diff in means", order = c("Intervention", "Control"))

# Simulate the diffs under the null hypothesis

Improvement.in.Sdep.null.dist <- sleep.df %>%
specify(response = Improvement.in.Sdep, explanatory = Group) %>%
  hypothesize(null = "independence") %>%
  generate(reps = 1000, Group = "permute") %>%
  calculate(stat = "diff in means", order = c("Intervention", "Control"))

Improvement.in.Sdep.null.dist %>% 
    get_pvalue(obs_stat = samp.diff, direction = "two_sided")


```

```{r}

Improvement.in.Sdep.null.dist %>% 
  ggplot(aes(stat)) + 
  geom_histogram() +
  geom_vline(data=samp.diff, aes(xintercept = stat), colour="red")
  # geom_vline(aes(xintercept = 2.83), colour="red")
  ggtitle("Boxplot of Improvement in Sleep Deprivation by Group")
  xlab("Permuted Differences") +
    ylim(-3, 3)


```

# How often is the sample difference in means bigger than those generated by the 1000 permutations ?

```{r}

sum(samp.diff$stat > Improvement.in.Sdep.null.dist$stat )

```

```{r}

1 - (sum(samp.diff$stat > Improvement.in.Sdep.null.dist$stat )/1000)

```



```{r}

sleep.df %>%
specify(response = Improvement.in.Sdep, explanatory = Group) %>%
  hypothesize(null = "independence") %>%
  generate(reps = 1000, Group = "permute") %>%
  calculate(stat = "t", order = c("Intervention", "Control")) %>%
    visualize(method = "simulation") 

# Theoretical t distribution
# using t statistic

sleep.df %>%
specify(response = Improvement.in.Sdep, explanatory = Group) %>%
  hypothesize(null = "independence") %>%
#  generate(reps = 1000, Group = "permute") %>%
  calculate(stat = "t", order = c("Intervention", "Control")) %>%
    visualize(method = "theoretical")

# Overlay theoretical distribution on top of randomized t-statistics
sleep.df %>%
specify(response = Improvement.in.Sdep, explanatory = Group) %>%
  hypothesize(null = "independence") %>%
  generate(reps = 1000, Group = "permute") %>%
  calculate(stat = "t", order = c("Intervention", "Control")) %>%
    visualize(method = "both")


```
