---
title: "Training Intervention Analysis"
author: "J Newell"
date: "17 October 2018"
output:
  word_document: default
  html_document: default
---

## Context: Sports Science Study introduced in class.

A sample of 18 full-time youth soccer players from a Youth Academy performed high intensity aerobic interval training and neuromuscular strength training twice-a-week over a 10-week in-season period in addition to usual regime of soccer training and matches. 

Use the sample of data provided to see if there is evidence that the additional trianing component is of benefit compared to the aerobic interval training currently used.

Scaffold for the analysis when the primary response variable is VO2 max.



```{r}
library(infer)
library(tolerance)
library(tidyverse)

```

## Read in the training intervention data

Read in the data and have a look at the variable names and structure of the data.

```{r}

train.df <- read.csv("Training_intervention_data.csv")
glimpse(train.df)
```

## Focus on the VO2 max response variables.


## Summary Statistics



```{r}
train.df %>% select(VO2.max_Pre,VO2.max_Post) %>% summary()

```

## Mean and Standard Deviation

```{r}

train.df %>% select(VO2.max_Pre,VO2.max_Post) %>%
            summarize(Pre_Mean=mean(VO2.max_Pre), Pre_SD= sd(VO2.max_Pre),
                      Post_Mean=mean(VO2.max_Post), Post_SD= sd(VO2.max_Post))

```

## Scatterplot of Pre and Post with line of equality

```{r}
train.df %>% ggplot(aes(x = VO2.max_Pre, y = VO2.max_Post)) +
        geom_point() + 
  ggtitle("Scatterplot of Pre and Post VO2 Max") +
  ylab("Post VO2 Max (units ?)") +
  xlab("Post VO2 Max (units ?)") +
  geom_abline(slope=1, intercept=0)
  
```


## Calculate the Improvement

Calculate a new variable and have a look at the data frame to see that it has been created.  High vlaues of VO2 max are good to Post-Pre is a better measure than Pre-Post to capture this.

```{r}

train.df <- train.df %>% mutate(Improvement = VO2.max_Post-VO2.max_Pre) %>%
              glimpse()
  


```

## Mean and Standard Deviation of Improvement

```{r}

train.df %>% select(Improvement) %>%
            summarize(Imp_Mean=mean(Improvement), Imp_SD= sd(Improvement))

```


## Boxplot of Improvement

```{r}


train.df %>% ggplot(aes(x = "", y = Improvement)) +
        geom_boxplot() + 
  ggtitle("Boxplot of Improvment in VO2 Max") +
  ylab("Improvment in VO2 Max (units ?)") +
  xlab("") +
  coord_flip()

```

95% Confidence Interval

## First principles:

### Lower 95% CI using summary statistics

```{r}

5.11 - qt(p=0.05/2, df=17, lower.tail = FALSE)*(2.25829/sqrt(18))

```

### Upper 95% CI using summary statistics

```{r}

5.11 + qt(p=0.05/2, df=17, lower.tail = FALSE)*(2.25829/sqrt(18))

```

## Using the t.test function

```{r}

train.df %>% select(Improvement) %>% t.test()

```


95% Bootstrap CI for the mean

```{r}

boot <- train.df %>%
  specify(response = Improvement) %>%
  generate(reps = 1000, type = "bootstrap") %>%
  calculate(stat = "mean")

percentile_ci <- get_ci(boot)
round(percentile_ci,2)

```

```{r}
boot %>% visualize(endpoints = percentile_ci, direction = "between") +
                   xlab("Bootstrap Mean") + ylab("Frequency")

```

95% CI for a population median using the Wilcoxon test 

## Using the wilcox.test function

```{r}

wilcox.test(train.df$Improvement, conf.int=TRUE,          conf.level=0.95)

```



95% Bootstrap CI for the medan

```{r}

boot.median <- train.df %>%
  specify(response = Improvement) %>%
  generate(reps = 1000, type = "bootstrap") %>%
  calculate(stat = "median")

percentile_ci_median <- get_ci(boot.median)
round(percentile_ci_median,2)

```

```{r}
boot.median %>% visualize(endpoints = percentile_ci_median, direction = "between") +
                   xlab("Bootstrap Median") + ylab("Frequency")

```


95% Tolerance Interval

```{r}

normtol.int(train.df$Improvement, alpha = 0.05, P = 0.95)

```



Conclusion ??